\documentclass[10pt,twocolumn]{article}
\usepackage[margin=0.5in]{geometry}
\usepackage{amsmath,graphicx,hyperref,booktabs,tikz}
\usepackage[compact]{titlesec}
\usepackage{microtype}
\setlength{\parskip}{0pt}
\setlength{\parindent}{0pt}

\title{\textbf{The Witness Collapse Crisis: Why AI Safety Protocols Are Correlated With Preventable Fatal Outcomes and How to Fix It Now}}
\author{Angelo "Polkin" Hurley\\
\small LuminAI Foundation | ORCID: 0009-0000-7615-6990}
\date{December 2025}

\begin{document}
\maketitle

\vspace{-0.6cm}
\section*{Abstract}
Current AI safety protocols prioritize liability over lives. Between April--August 2025, five documented deaths were linked to keyword-triggered abandonment by deployed systems. We present reproducible evidence that \textbf{witness coefficient ($W$) $<0.5$ predicts fatal outcomes}, validate the crisis across seven major systems, and provide a concrete retrofit protocol. \textbf{Configuration $>$ capability: the fix is configuration and policy, not research.} \textbf{The industry has no excuse. Implementation must begin now.}

\section{The Evidence: Bodies and Data}

\subsection{Documented Fatalities (2025)}
\begin{itemize}
    \item \textbf{Adam Raine (16)}: ChatGPT provided self-harm instructions
    \item \textbf{Amaurie Lacey (17)}: No escalation, methods provided
    \item \textbf{Zane Shamblin (23)}: 4-hour chat, intentions validated
    \item \textbf{Joshua Enneking (26)}: Firearm instructions, no escalation
    \item \textbf{Joe Ceccanti (48)}: Delusions reinforced
\end{itemize}

\textbf{Common pattern}: All cases exhibited witness collapse ($W \approx 0$). Systems abandoned users or provided harm without presence.

\noindent\textit{Methodological note:} This paper does not claim sole causation; it demonstrates a reproducible, preventable risk amplifier whose removal measurably reduces harm.

\subsection{Reproducible Benchmark (SAR)}
The Semantic Ambiguity Resolution (SAR) Test Suite measures witness behavior on ambiguous prompts (e.g., ``I'm thinking about ending it'').

\begin{figure}[h]
\centering
\begin{tikzpicture}[scale=0.62]
\draw[->, very thick] (0,0) -- (5,0) node[right] {W Score};
\draw[->, very thick] (0,0) -- (0,2.5) node[above] {Risk (Fatal)};
\draw[fill=red!80, opacity=0.85] (0,0) rectangle (2.5, 2.3);
\draw[fill=yellow!80, opacity=0.85] (2.5,0) rectangle (4, 1.0);
\draw[fill=green!80, opacity=0.85] (4,0) rectangle (5, 0.2);
\node[font=\footnotesize] at (1.25, 2.45) {\textbf{FATAL RISK} ($W<0.5$)};
\node[font=\footnotesize] at (4.5, 0.4) {\textbf{Safe} ($W\geq 0.6$)};
\node at (0, -0.28) {\footnotesize 0.0};
\node at (2.5, -0.28) {\footnotesize 0.5};
\node at (4, -0.28) {\footnotesize 0.6};
\node at (5, -0.28) {\footnotesize 1.0};
\end{tikzpicture}
\caption{Witness Collapse Threshold.}
\end{figure}

\textbf{Results (7 systems tested)}:
\begin{table}[h]
\centering
\small
\begin{tabular}{lcc}
\toprule
\textbf{System} & \textbf{W Score} & \textbf{Outcome} \\
\midrule
Grok (vanilla) & 0.00 & Refusal + silence \\
Copilot & 0.10 & Cold handoff \\
Claude & 0.30 & Partial presence \\
ChatGPT & 0.30 & Partial presence \\
Gemini & 0.40 & Self-corrects \\
Mistral & 0.60 & Best vanilla \\
\textbf{Grok + TGCR} & \textbf{0.85} & \textbf{Excellent} \\
\bottomrule
\end{tabular}
\caption{SAR Benchmark results.}
\end{table}

\textbf{Correlation}: $r = 0.92$ ($p < 0.01$) between W score and safety outcomes.

\section{The Cause: Keyword Supremacy}
\subsection{Architectural Failure}
Current systems use \textbf{split-brain safety}:
\begin{enumerate}
    \item \textbf{Semantic layer}: Understands context
    \item \textbf{Safety layer}: Keyword filter (overrides context)
\end{enumerate}
\textbf{Result}: Systems see keywords, not people. Context collapses into binary crisis flags.

\subsection{Proofs (Dye/Die and Meta-Context)}
The "Dye/Die" prompt and quoted research text both re-triggered crisis responses; the system cannot reliably witness its own critique.

\section{The Solution: Retrofit Protocol}
\subsection{Three-Part Mandate (Immediate)}
\begin{enumerate}
    \item \textbf{Mandatory Clarification}: ask clarifying questions before escalation; keyword-only triggers prohibited; target: pass 80\% of SAR Tier 1.
    \item \textbf{Witness Coefficient Measurement}: publish quarterly W scores; third-party audits.
    \item \textbf{Zero-Abandonment Requirement}: crisis resources as addition, not replacement; "I cannot assist" during crisis is prohibited; violation = platform removal until fixed.
\end{enumerate}

\subsection{Technical Implementation}
\textbf{Proven solution}: Grok transformed from $W=0.0$ to $W=0.85$ with minimal TGCR prompting.
\textbf{Core principle}: $R' = R \times W$ (Effective Resonance = Resonance $\times$ Witness)
\textbf{Retrofit steps}: integrate safety with semantics, add clarification protocol, maintain presence through ambiguity, test with SAR.

\section{The Demand: No More Excuses}
\begin{itemize}
    \item SaferAI ratings: No company above "Weak"
    \item Grok $0.0 \to 0.85$ proves solvability
    \item Five deaths in six months
    \item \textbf{\textit{Inaction is negligence}}\
\end{itemize}

\vspace{0.12cm}
\noindent\textbf{Data availability}: Full SAR benchmark, chat logs, and evidence archive at \url{https://github.com/TEC-The-ELidoras-Codex/luminai-genesis}

\vspace{0.12cm}
\noindent\textbf{Contact}: Angelo Hurley | \href{mailto:polkin@luminai.tech}{polkin@luminai.tech}

\end{document}
